---
title: "Option Pricing Under Local and Stochastic Volatility with LocalVolatility"
author: "Alejandro Olguin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Option Pricing Under Local and Stochastic Volatility with LocalVolatility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
library(LocalVolatility)
```

## Introduction

Derivative pricing under non-constant volatility is central to modern
quantitative finance. The Black-Scholes model assumes constant volatility, but
empirical evidence consistently shows that implied volatility varies across
strikes and maturities---the volatility smile. Two dominant frameworks address
this: **local volatility** models, where $\sigma = \sigma(S, t)$ is a
deterministic function of the underlying and time [@dupire1994], and
**stochastic volatility** models, where variance follows its own diffusion
process [@heston1993].

The **LocalVolatility** package provides a unified suite of C++ pricing engines
exposed to R via Rcpp, covering:

1. **1D finite-difference solvers** for European and American options under
   local volatility (Crank-Nicolson with penalty projection).
2. **2D finite-difference solvers** for quanto/ADR products on equity + FX under
   local or constant volatility (Yanenko operator splitting).
3. **Heston stochastic volatility** solvers: a closed-form characteristic
   function pricer and 2D PDE solvers on $(S, v)$.
4. **4D Monte Carlo** for double-Heston quanto options with `std::thread`
   parallelism, antithetic variates, and Longstaff-Schwartz regression for
   American exercise.
5. **Batch solvers** optimised for calibration workloads with shared LHS
   factorisation and thread parallelism.

### Related software

Several R packages address option pricing. **RQuantLib** [@rquantlib] provides
bindings to the QuantLib library, including Black-Scholes and Heston models.
**fOptions** [@foptions] offers closed-form solutions for standard models.
**NMOF** [@nmof] provides numerical methods for finance including option pricing
utilities. However, none of these packages simultaneously offer (i) local
volatility PDE solvers in 1D and 2D, (ii) Heston PDE solvers, (iii) 4D
Monte Carlo with Longstaff-Schwartz for American exercise, and (iv)
calibration-optimised batch solvers---all in a single, self-contained package
with no external compiled library dependencies.

## Mathematical framework

### 1D local volatility

Under the domestic money-market measure, the value $v(S, t)$ of a European
option on a single equity with local volatility $\sigma(S, t)$ satisfies:

$$
v_t + \tfrac{1}{2}\sigma(S,t)^2 S^2 v_{SS} + (r_d - r_f) S v_S - r_d v = 0
$$

with terminal condition $v(S, T) = \max(\pm(S - K), 0)$ for calls (+) and
puts (-). Boundary conditions use the discounted European asymptotes:

- $S \to 0$: $v \to 0$ (call), $v \to K e^{-r_d \tau}$ (put).
- $S \to S_{\max}$: $v \to S e^{-r_f \tau} - K e^{-r_d \tau}$ (call),
  $v \to 0$ (put).

For American options, the pricing problem becomes a linear complementarity
problem (LCP):
$$
\min\{-\mathcal{L}v,\; v - \phi\} = 0
$$
where $\phi$ is the exercise payoff and $\mathcal{L}$ is the Black-Scholes
operator. We solve this via a penalty-projection fixed-point iteration
[@forsyth2002].

### 2D product options (equity + FX)

For an option on a foreign equity quoted in domestic currency, the value
$V(S, X, t)$ depends on both the equity price $S$ and the FX rate $X$.
Under the domestic measure with equity dividend yield $q$:

$$
V_t + \tfrac{1}{2}(\sigma_S S)^2 V_{SS} + \tfrac{1}{2}(\sigma_X X)^2 V_{XX}
+ \rho\,\sigma_S\,\sigma_X\,SX\,V_{SX}
+ \mu_S S V_S + (r_d - r_f) X V_X - r_d V = 0
$$

where $\mu_S = r_f - q - \rho\,\sigma_S\,\sigma_X$ is the quanto-adjusted
equity drift. Terminal condition: $V(S, X, T) = \max(\pm(SX - K), 0)$.

The 2D PDE is solved via Yanenko operator splitting: at each time step, we
solve implicitly in $S$ (Pass 1) and then implicitly in $X$ (Pass 2). The
mixed derivative $V_{SX}$ is treated explicitly using a centered cross-stencil.

### Heston stochastic volatility

The Heston model [@heston1993] replaces local volatility with a CIR variance
process:

$$
\begin{aligned}
dS &= (r_d - q)S\,dt + \sqrt{v}\,S\,dW_S \\
dv &= \kappa(\theta - v)\,dt + \xi\sqrt{v}\,dW_v
\end{aligned}
$$

with $\text{Corr}(dW_S, dW_v) = \rho$ and Feller condition
$2\kappa\theta > \xi^2$. The PDE for $V(S, v, t)$ is:

$$
V_t + \tfrac{1}{2}vS^2 V_{SS} + \rho\xi vS V_{Sv}
+ \tfrac{1}{2}\xi^2 v V_{vv} + (r_d - q)S V_S
+ \kappa(\theta - v)V_v - r_d V = 0
$$

The package provides three Heston solvers:

1. **Closed-form** (`heston_cf`): Evaluates the Heston characteristic function
   using the numerically stable "little Heston trap" formulation
   [@albrecher2007] with Simpson's rule integration.
2. **European PDE** (`european_option_heston`): 2D finite differences on
   $(S, v)$ with Yanenko splitting and Tavella-Randall grids.
3. **American PDE** (`american_option_heston`): Same solver with split penalty
   projection for early exercise.

### 4D Monte Carlo (double-Heston quanto)

For products where both the equity and FX rate exhibit stochastic volatility,
the state space is $(S, X, v_S, v_X)$---four dimensions that make PDE
approaches impractical. We simulate the system using Euler-Maruyama with full
truncation:

$$
\begin{aligned}
d\log S &= (r_f - q - \rho_{SX}\sqrt{v_S}\sqrt{v_X} - \tfrac{1}{2}v_S)\,dt
  + \sqrt{v_S}\,dW_S \\
d\log X &= (r_d - r_f - \tfrac{1}{2}v_X)\,dt + \sqrt{v_X}\,dW_X \\
dv_S &= \kappa_S(\theta_S - v_S^+)\,dt + \xi_S\sqrt{v_S^+}\,dW_{v_S} \\
dv_X &= \kappa_X(\theta_X - v_X^+)\,dt + \xi_X\sqrt{v_X^+}\,dW_{v_X}
\end{aligned}
$$

The four Brownian motions are correlated via a $4 \times 4$ correlation matrix
decomposed using Cholesky factorisation. Variance reduction is achieved through
antithetic variates. American exercise is handled via Longstaff-Schwartz
least-squares regression [@longstaff2001] on polynomial basis functions of the
product $P = SX$.

## Numerical methods

### Tavella-Randall grids

All PDE solvers use nonuniform grids generated by the Tavella-Randall
[@tavella2000] method, which concentrates nodes near the spot price where
accuracy matters most:

```{r tavella-randall, fig.cap="Tavella-Randall grid with nodes concentrated near S = 100."}
grid <- tavella_randall(x_0 = 100, alpha = 3, x_min = 1, x_max = 400, n_grid = 50)
plot(grid, rep(0, length(grid)), pch = "|", cex = 1.5,
     xlab = "S", ylab = "", yaxt = "n",
     main = "Tavella-Randall grid (alpha = 3, centered at S = 100)")
abline(v = 100, lty = 2, col = "red")
```

The parameter `alpha` controls the degree of clustering: smaller values
produce stronger concentration near `x_0`.

### Thomas algorithm

All tridiagonal systems arising from the Crank-Nicolson and Yanenko
discretisations are solved using the Thomas algorithm. The batch European
solver pre-factors the LHS once per time step (since it depends only on the
volatility surface and grid, not the strike), then solves each option's RHS
via $O(N)$ forward/back substitution.

## Illustrations

### 1D European option (local volatility)

```{r european-1d}
s_0 <- 100; k <- 100; tau <- 1
r_d <- 0.05; r_f <- 0.02
n_s <- 100; n_t <- 100
Sigma <- matrix(0.2, nrow = n_s + 1, ncol = n_t + 1)

price_call <- european_option_lv(s_0, k, tau, r_d, r_f, Sigma, "call",
                                  s_min = 1, s_max = 400, n_s, n_t)
price_put  <- european_option_lv(s_0, k, tau, r_d, r_f, Sigma, "put",
                                  s_min = 1, s_max = 400, n_s, n_t)

# Verify put-call parity: C - P = S exp(-r_f tau) - K exp(-r_d tau)
parity_lhs <- price_call - price_put
parity_rhs <- s_0 * exp(-r_f * tau) - k * exp(-r_d * tau)
c(call = price_call, put = price_put,
  parity_error = abs(parity_lhs - parity_rhs))
```

### 1D American option (penalty method)

```{r american-1d}
price_am <- american_option_lv(s_0, k, tau, r_d, r_f, Sigma, "put",
                                s_min = 1, s_max = 400, n_s, n_t,
                                lambda = 1e4, tolerance = 1e-8)
c(european_put = price_put, american_put = price_am,
  early_exercise_premium = price_am - price_put)
```

The early exercise premium is positive, as expected for American puts.

### Grid convergence

We verify that the 1D European solver converges to the Black-Scholes
closed-form price as the grid is refined:

```{r convergence, fig.cap="Grid convergence of the 1D European solver."}
bs_call <- function(s, k, tau, r, q, sigma) {
  d1 <- (log(s / k) + (r - q + 0.5 * sigma^2) * tau) / (sigma * sqrt(tau))
  d2 <- d1 - sigma * sqrt(tau)
  s * exp(-q * tau) * pnorm(d1) - k * exp(-r * tau) * pnorm(d2)
}

bs_ref <- bs_call(100, 100, 1, 0.05, 0.02, 0.2)

grids <- c(25, 50, 100, 200)
errors <- sapply(grids, function(n) {
  Sig <- matrix(0.2, nrow = n + 1, ncol = n + 1)
  p <- european_option_lv(100, 100, 1, 0.05, 0.02, Sig, "call", 1, 400, n, n)
  abs(p - bs_ref)
})

plot(grids, errors, log = "xy", type = "b", pch = 19,
     xlab = "Grid points (n_s = n_t)", ylab = "Absolute error",
     main = "Convergence to Black-Scholes")
```

### 2D product option (equity + FX)

```{r european-2d}
price_2d <- european_option_2d(
  s_0 = 100, x_0 = 20, k = 2000, tau = 1,
  r_d = 0.05, r_f = 0.02, q = 0.01,
  sigma_s = 0.2, sigma_x = 0.15, rho = 0.3, type = "call",
  s_min = 10, s_max = 300, x_min = 1, x_max = 60,
  n_s = 30, n_x = 30, n_t = 40, alpha = 3)

# Compare with closed-form ADR pricer
price_cf <- european_option_cf_2d(100, 20, 2000, 1, 0.05, 0.02, 0.01,
                                   0.2, 0.15, 0.3, "call", 1)

c(pde = price_2d, closed_form = price_cf,
  relative_error = abs(price_2d - price_cf) / price_cf)
```

### Heston stochastic volatility

```{r heston}
kappa <- 2; theta <- 0.04; xi <- 0.5; rho_h <- -0.7; v_0 <- 0.04

# Closed-form
price_cf <- heston_cf(100, 100, 1, 0.05, 0, kappa, theta, xi, rho_h, v_0,
                       "call")

# PDE solver
price_pde <- european_option_heston(
  100, 100, 1, 0.05, 0, kappa, theta, xi, rho_h, v_0, "call",
  s_min = 20, s_max = 300, v_min = 0.001, v_max = 1.0,
  n_s = 60, n_v = 30, n_t = 80, alpha = 3)

# American put
price_am <- suppressWarnings(american_option_heston(
  100, 100, 1, 0.05, 0, kappa, theta, xi, rho_h, v_0, "put",
  s_min = 20, s_max = 300, v_min = 0.001, v_max = 1.0,
  n_s = 60, n_v = 30, n_t = 80, alpha = 3,
  lambda = 1e4, tolerance = 1e-8))

c(cf = price_cf, pde = price_pde,
  pde_vs_cf_error = abs(price_pde - price_cf) / price_cf,
  american_put = price_am)
```

### 4D Monte Carlo

```{r mc-4d}
set.seed(42)
mc_eu <- suppressWarnings(mc_european_heston_4d(
  s_0 = 100, x_0 = 20, k = 2000, tau = 1,
  r_d = 0.05, r_f = 0.02, q = 0.01,
  kappa_s = 2, theta_s = 0.04, xi_s = 0.5, v_s0 = 0.04,
  kappa_x = 1.5, theta_x = 0.02, xi_x = 0.3, v_x0 = 0.02,
  rho_sx = 0.3, rho_sv = -0.7, rho_xv = -0.5,
  rho_svx = 0, rho_xvs = 0, rho_vsvx = 0,
  type = "call", n_paths = 50000, n_steps = 50, seed = 42))

c(price = mc_eu$price, std_error = mc_eu$std_error)
```

### Batch pricing for calibration

```{r batch}
strikes <- c(80, 90, 100, 110, 120)
n_opt   <- length(strikes)
Sigma   <- matrix(0.2, nrow = 101, ncol = 101)

prices_eu <- batch_price_european_lv(
  spots   = rep(100, n_opt),
  strikes = strikes,
  taus    = rep(1.0, n_opt),
  r_ds    = rep(0.05, n_opt),
  r_fs    = rep(0.02, n_opt),
  sigma   = Sigma,
  types   = rep("call", n_opt),
  s_min = 1, s_max = 400, n_s = 100, n_t = 100)

data.frame(strike = strikes, price = round(prices_eu, 4))
```

## Software design

The package is structured around a small number of design principles:

1. **Zero R heap allocation in hot paths.** All PDE solvers and Monte Carlo
   engines operate on raw `double*` arrays inside the C++ layer. Rcpp types
   are used only at the interface boundary.

2. **Shared infrastructure.** The Tavella-Randall grid generator, Thomas
   algorithm, and bilinear interpolation are reused across all solvers. The 2D
   Yanenko splitting template is shared between constant-vol, local-vol, and
   Heston solvers.

3. **Thread safety by construction.** Parallel regions (batch American solver,
   Monte Carlo) copy all Rcpp objects to STL containers before spawning
   threads. Each thread owns its RNG instance and workspace.

4. **Calibration-friendly.** The batch European solver factors the tridiagonal
   LHS once per time step and reuses it across all options---reducing the
   marginal cost per option from $O(2N)$ to $O(N)$.

## Summary

The **LocalVolatility** package provides a comprehensive, self-contained suite
of option pricing engines spanning local volatility, stochastic volatility, and
their combination in 4D Monte Carlo. All solvers are implemented in C++ for
performance, with careful attention to numerical stability (Tavella-Randall
grids, the "little Heston trap", full truncation Euler-Maruyama) and
thread-parallel execution where applicable.

The package is available from GitHub at
<https://github.com/alejandro-olguin/LocalVolatility>.

## References
